Early Observations of the Type Ia Supernova \abc:\\ XXX NEED SOMETHING CLEVER TO SAY HERE XXX

### 2017-08-04 doc/ms.tex
Update the manuscript to reflect Brad's comments

\title{Let There Be Light: \\ Early Observations of the Normal Type Ia Supernova \abc\ \\Show No Evidence for a Dark Period}

@@

Early observations of Type Ia supernovae (SNe) provide a unique probe of their progenitor systems and explosion physics. Here, we report the intermediate Palomar Transient Factory (iPTF) discovery of an extraordinarily young SN Ia, \abc. The initial detection was made only $0.15\pm^{0.07}_{0.15}$~d after SN first light. In the $\sim$24 hr after discovery, \abc\ rose by $\sim$2 mag, following a near-linear rise in flux for $\sim$4 d. Assuming the photospheric velocity follows a $t^{-0.22}$ power law, the time of explosion for \abc\ is approximately equal to the time of first light. Strong \ion{C}{2} absorption is detected in the early spectra of \abc, before disappearing after $\sim$7~d. Unlike SN\,2011fe, the $(B-V)_0$ colors of \abc\ are blue and nearly constant in the days after explosion. We show that the fast, near-linear rise in flux for \abc\ cannot be explained by either SN shock breakout and the associated, subsequent cooling or the collision of the SN ejecta with a stellar companion. Instead, we argue that the early characteristics of \abc, including: (i) the rapid, near-linear rise, (ii) the non-evolving blue colors, and (iii) the strong lines from ionized carbon in the earliest spectra, are the result of either vigorous mixing of radioactive $^{56}$Ni in the SN ejecta, or the interaction of the ejecta with diffuse material, or a combination of the two. In the next few years, dozens of very young \textit{normal} SNe Ia will be discovered, and observations similar to those presented here will constrain the white dwarf explosion mechanism.

@@

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{methods: observational --- supernovae: individual (\abc)}

@@

During the spring of 2016, the iPTF survey observed the field of \abc\ every night in either the $g_\mathrm{PTF}$- or 
$R_\mathrm{PTF}$-band.\footnote{P48 observations of \abc\ are reported in the $g_\mathrm{PTF}$ and $R_\mathrm{PTF}$ filters throughout, which are similar to the SDSS $g'$ and $r'$ filters, respectively (see \citealt{2012PASP..124..854O} for details on PTF calibration). The correction from the $g_\mathrm{PTF}$ and $R_\mathrm{PTF}$ filters to SDSS $g'$ and $r'$ requires knowledge of the intrinsic source color (see Eqns.~1 and 2 in \citealt{2012PASP..124..854O}). The spectral diversity of SNe Ia in the days after explosion is poorly constrained, and as a result the color terms for \abc\ at these epochs are unknown. We proceed by assuming the $g_\mathrm{PTF}$ and $R_\mathrm{PTF}$ calibration is on the AB system, which strictly speaking is incorrect, but this does not fundamentally alter any of our conclusions.} 
Survey observations were conducted with the
CFH12K camera \citep{2000SPIE.3965...58S} on the Palomar Observatory 
48-inch telescope (P48; \citealt{2009PASP..121.1395L}). Images were processed by the IPAC image
subtraction and discovery pipeline which subtracts off the background
galaxy light with stacked pre-SN images and performs forced
point-spread-function (PSF) photometry at the location of the SN 
\citep{2017PASP..129a4002M}. The
photometry is then calibrated to the PTF photometric catalog
\citep{2012PASP..124..854O}.

@@

The $g$-band discovery and follow-up data of \abc\ are 
illustrated in Figure~\ref{fig:lightcurve}.  The photometry is shown in the AB system with a zero point of 
$3631$\,Jy. As previously noted the color terms necessary to convert $g_\mathrm{PTF}$ to the AB system are unknown and assumed to be zero.

@@

\section{Reddening, Classification and Host Galaxy}
\label{sec:usual_staff}

@@

A detailed study of the reddening towards \abc\ is presented in a 
companion paper (Ferretti et al.\ 2017, submitted). Briefly, the 
foreground Galactic extinction toward \abc\
is $E(B-V) = 0.0279 \, \mathrm{mag}$ \citep{2011ApJ...737..103S}. The high-resolution spectra presented in Ferretti et al.\ (2017) show multiple absorption components for both the \ion{Ca}{2}\,H$+$K and \ion{Na}{1}\,D doublets. While the equivalent width (EW) of these lines is quite large, implying a large amount of extinction (e.g., \citealt{2012MNRAS.426.1465P}), Ferretti et al.\ compare the evolution of \abc\ to the well-observed normal type Ia SN\,2011fe and find evidence for only a small amount of extinction. The empirical relation between the EW of \ion{Na}{1}\,D and extinction is known to have a large scatter, and \citet{2013ApJ...779...38P} have shown that \ion{Na}{1}\,D absorption is a poor tracer of reddening in Type Ia SNe. Thus, we instead adopt $E(B-V) = 0.05 \, \mathrm{mag}$ as the local extinction for \abc\ (Ferretti et al.\ 2017). For the remainder of our analysis we therefore adopt a total, Galactic $+$ host galaxy, line-of-sight extinction of $E(B-V) = 0.0779 \, \mathrm{mag}$.

@@

Using the Supernova Identification (\texttt{SNID}; \citealt{2007ApJ...666.1024B}) package, 
we find the low-resolution spectrum of \abc\ at $+18.8$ d is best matched by normal SNe Ia. Several characteristic features of a SN
Ia, such as \ion{Si}{2}, \ion{S}{2}, can be easily identified in the
spectra of \abc\ (Figure \ref{fig:spec_seq}).

@@

For convenience, in the following sections, we define the best-fit
value $\textrm{MJD}_{max}=57499.54$ as the time of $B$-band maximum, $T_{B_\mathrm{max}}$, which we also adopt as phase $t=0$.

After establishing \abc\ as a normal SN Ia, we use the latest
calibration \citep{2014A&A...568A..22B} of the Phillips relation
\citep{1993ApJ...413L.105P} using $m^*_{B}$, $x_1$ and $c$ to derive 
a distance modulus $\mu = 34.88 \pm 0.10 \,\textrm{mag}$ to the SN, 
provided that the host galaxy of \abc\ has a stellar mass less than
$10^{10}\sm$. We note that a more massive host galaxy would result 
in a larger inferred distance modulus that is nevertheless 
consistent within the uncertainties.

The location of \abc\ is spatially coincident with a tidal tail of
galaxy NGC\,5221. \citet{2007A&A...465...71T} derived a distance
modulus of $35.0\pm0.4\,\textrm{mag}$ to NGC\,5221 from the 
Tully-Fisher relation, which is consistent with that of \abc.

@@

As our observations include especially early observations of \abc, there are upper limits $\sim$1 d prior to the discovery epoch, we model the early flux from \abc\ as a power law, but allow the power-lax index to vary, as opposed to fixing it at 2, to account for potential variations in the photospheric temperature during expansion:
\begin{equation}
  \label{eq:broken_power_law}
  f(t) \left\{
    \begin{array}{ll}
      = 0,\ \textrm{when}\ t<t_0 \\
      \propto (t-t_0)^{\alpha},\ \textrm{when}\ t>t_0
    \end{array}
  \right.\ ,
\end{equation}
where $t_0$ is the time of first light, $\alpha$ is the power-law index, and $t$ is measured in the rest-frame of the SN. To determine $t_0$ and $\alpha$ we fit the earliest observations of \abc. Due to slight variations in the passbands, we fit the model to the relative flux measured in the $g_\mathrm{PTF}$-band, which is the only filter with observations prior to first light, a necessity for constraining $t_0$.

@@

Comparing our estimates for $t_{exp}$ and $t_0$ (left panel of 
Figure \ref{fig:velocity_t_exp}), we find that 
$t_0\lessapprox t_{exp}$.
Since physical causality requires $t_{exp}<t_0$, we draw the qualitative
conclusion that $t_0\simeq t_{exp}$, which is consistent to within the uncertainties. Again, we caution that this derivation of $t_\mathrm{exp}$ relies on the assumption $v \propto t^{-0.22}$, which may not be valid for all SNe Ia.

@@

To examine the possibility of a SN-companion signature in the early
light curve of \abc, we employ the \citet{2010ApJ...708.1025K}
model and assume canonical values for the ejecta mass, $1.4\,\sm$, expansion velocity,
$10^{4}\,\textrm{km}\,\textrm{s}^{-1}$, and a constant opacity, 
$0.2\,\textrm{cm}^2\,\textrm{g}^{-1}$. We calculate the expected 
$g_\mathrm{PTF}$ brightness of an ejecta-companion collision at the 
distance of \abc\ behind a total redenning of $E(B-V) = 0.0779 \, \mathrm{mag}$ using the parameterized equations in 
\citet{2012ApJ...749...18B}. If we assume the binary is aligned with the optimal orientation relative to the line of sight, a binary separation of $a \approx 2 \times 10^{13}\, \mathrm{cm}$ is needed to explain the initial detection of \abc, as shown in Figure~\ref{fig:SN-companion}. This model does not, however, match the $g_\mathrm{PTF}$ evolution for $t > t_0 + 1 \, \mathrm{d}$ (though it is possible that the SN photosphere dominates the companion-interaction signature at this phase). Furthermore, this model significantly overpredicts the observed flux in the \textit{Swift}/$UVW1$ filter $\sim$3.1~d after $t_0$. As such, we do not favor such a model as the explanation for the early emission from \abc.

@@

The Intermediate Palomar Transient Factory project is a scientific collaboration among the California Institute of Technology, Los Alamos National Laboratory, the University of Wisconsin, Milwaukee, the Oskar Klein Center, the Weizmann Institute of Science, the TANGO Program of the University System of Taiwan, and the Kavli Institute for the Physics and Mathematics of the Universe. This work was supported by the GROWTH project funded by the National Science Foundation under Grant No 1545949. Part of this research was carried out at the Jet Propulsion Laboratory, California Institute of Technology, under a contract with the NASA. This work makes use of observations from the LCO network.

@@

\software{\texttt{PTFIDE} \citep{Masci2017PASP}, \texttt{SALT2}, \texttt{SNID}, \texttt{UVOTSOURCE}}.

>>>>

\title{Early Observations of the Type Ia Supernova \abc:\\
XXX NEED SOMETHING CLEVER TO SAY HERE XXX}

@@

Early observations of Type Ia supernovae (SNe) provide a unique probe of their progenitor systems and explosion physics. Here, we report the intermediate Palomar Transient Factory (iPTF) discovery of an extraordinarily young SN Ia, \abc. The initial detection was made only $0.15\pm^{0.07}_{0.15}$~d after SN first light, the time when radiation from $^{56}$Ni decay has diffused to the SN photosphere. In the $\sim$24 hr after discovery, \abc\ rose by $\sim$2 mag, following a near-linear rise in flux for $\sim$4 d. 
% Assuming the photospheric velocity follows a $t^{-0.22}$ power law, the time of explosion for \abc\ is approximately equal to the time of first light.
Strong \ion{C}{2} absorption is detected in the early spectra of \abc, before disappearing after $\sim$7~d. Unlike SN\,2011fe, the $(B-V)_0$ colors of \abc\ are blue and nearly constant in the days after explosion. We show that the fast, near-linear rise in flux for \abc\ cannot be explained by either SN shock breakout and the associated, subsequent cooling or the collision of the SN ejecta with a stellar companion. Instead, we argue that the early characteristics of \abc, including: (i) the rapid, near-linear rise, (ii) the non-evolving blue colors, and (iii) the strong lines from ionized carbon in the earliest spectra, are the result of either vigorous mixing of radioactive $^{56}$Ni in the SN ejecta, or the interaction of the ejecta with diffuse material, or a combination of the two. In the next few years, dozens of very young \textit{normal} SNe Ia will be discovered, and observations similar to those presented here will constrain the white dwarf explosion mechanism.

@@

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
\keywords{methods: observational --- supernovae: general --- supernovae: individual (\abc)}

@@

During the spring of 2016, the iPTF survey observed the field of \abc\ every night in either the $g_\mathrm{PTF}$- or 
$R_\mathrm{PTF}$-band.\footnote{P48 observations of \abc\ are reported in the $g_\mathrm{PTF}$ and $R_\mathrm{PTF}$ filters throughout, which are similar to the SDSS $g'$ and $r'$ filters, respectively (see \citealt{2012PASP..124..854O} for details on PTF calibration). The correction from the $g_\mathrm{PTF}$ and $R_\mathrm{PTF}$ filters to SDSS $g'$ and $r'$ requires knowledge of the intrinsic source color (see Eqns.~1 and 2 in \citealt{2012PASP..124..854O}). The spectral diversity of SNe Ia in the days after explosion is poorly constrained, and as a result the color terms for \abc\ at these epochs are unknown. We proceed by assuming the $g_\mathrm{PTF}$ and $R_\mathrm{PTF}$ calibration is on the AB system, which strictly speaking is incorrect, but this does not fundamentally alter any of our conclusions.} 
Survey observations were conducted with the
CFH12K camera \citep{2008SPIE.7014E..4YR} on the Palomar Observatory 
48-inch telescope (P48; \citealt{2009PASP..121.1395L}). Images were processed by the IPAC image
subtraction and discovery pipeline which subtracts off the background
galaxy light with stacked pre-SN images and performs forced
point-spread-function (PSF) photometry at the location of the SN 
\citep{2017PASP..129a4002M}. The
photometry is then calibrated to the PTF photometric catalog
\citep{2012PASP..124..854O}.

@@

The $g$-band discovery and follow-up data of \abc\ are 
illustrated in Figure~\ref{fig:lightcurve}.  The photometry is shown in the AB system. As previously noted the color terms necessary to convert $g_\mathrm{PTF}$ to the AB system are unknown and assumed to be zero.

@@

\section{Host Galaxy, Reddening, and Classification}
\label{sec:usual_staff}

The location of \abc\ is spatially coincident with a tidal tail of
galaxy NGC\,5221. \citet{2007A&A...465...71T} derived a distance
modulus of $35.0\pm0.4\,\textrm{mag}$ to NGC\,5221 from the 
Tully-Fisher relation, which is consistent with what we later derive from the SN light curve (see \S\ref{sec:classification}).

@@

A detailed study of the reddening towards \abc\ is presented in a 
companion paper (Ferretti et al.\ 2017, submitted). Briefly, the 
foreground Galactic extinction toward \abc\
is $E(B-V) = 0.0279 \, \mathrm{mag}$ \citep{2011ApJ...737..103S}. The high-resolution spectra presented in Ferretti et al.\ (2017) show multiple absorption components for both the \ion{Ca}{2}\,H$+$K and \ion{Na}{1}\,D doublets. While the equivalent width (EW) of these lines is quite large, implying a large amount of extinction (e.g., \citealt{2012MNRAS.426.1465P}), Ferretti et al.\ compare the evolution of \abc\ to the well-observed normal type Ia SN\,2011fe and find evidence for only a small amount of extinction. The empirical relation between the EW of \ion{Na}{1}\,D and extinction is known to have a large scatter, and \citet{2013ApJ...779...38P} have shown that \ion{Na}{1}\,D absorption is a poor tracer of reddening in Type Ia SNe. Thus, we instead adopt $E(B-V) = 0.05 \, \mathrm{mag}$ as the local extinction for \abc\ (Ferretti et al.\ 2017). For the remainder of our analysis we therefore adopt a total, Galactic $+$ host galaxy, line-of-sight extinction of $E(B-V) = 0.08 \, \mathrm{mag}$.

@@

Using the SuperNova IDentification (\texttt{SNID}; \citealt{2007ApJ...666.1024B}) package, 
we find the low-resolution spectrum of \abc\ at $+18.8$ d is best matched by normal SNe Ia. Several characteristic features of a SN
Ia, such as \ion{Si}{2}, \ion{S}{2}, can be easily identified in the
spectra of \abc\ (Figure \ref{fig:spec_seq}).

@@

In the following sections, we define the best-fit
value $\textrm{MJD}_{max}=57499.54$ as the time of $B$-band maximum, $T_{B_\mathrm{max}}$, which we also adopt as phase $t=0$.

After establishing \abc\ as a normal SN Ia, we use the latest
calibration \citep{2014A&A...568A..22B} of the Phillips relation
\citep{1993ApJ...413L.105P} using $m^*_{B}$, $x_1$ and $c$ to derive a distance modulus $\mu = 34.89 \pm 0.10 \,\textrm{mag}$ to the SN, provided that the host galaxy of \abc\ has a stellar mass less than $10^{10}\sm$. We note that a more massive host galaxy would result in a larger inferred distance modulus that is nevertheless consistent within the uncertainties. For the following analysis we adopt a distance modulus $\mu = 34.89 \pm 0.10 \,\textrm{mag}$ for \abc.

@@

As our observations include especially early observations of \abc, there are upper limits $\sim$1 d prior to the discovery epoch, we model the early flux from \abc\ as a power law, but allow the power-lax index to vary, as opposed to fixing it at 2, to account for potential variations in the photospheric temperature during expansion:
\begin{equation}
  \label{eq:broken_power_law}
  f(t) \left\{
    \begin{array}{ll}
      = 0,\ \textrm{when}\ t<t_0 \\
      \propto (t-t_0)^{\alpha},\ \textrm{when}\ t>t_0
    \end{array}
  \right.\ ,
\end{equation}
where $t_0$ is the time of first light, $\alpha$ is the power-law index, and $t$ is measured in the rest-frame of the SN. To determine $t_0$ and $\alpha$ we fit the earliest observations of \abc. Due to slight variations in the passbands, we fit the model only to the relative flux measured in the $g_\mathrm{PTF}$-band, which is the only filter with observations prior to first light, a necessity for constraining $t_0$.

@@

Comparing our estimates for $t_{exp}$ and $t_0$ (left panel of 
Figure \ref{fig:velocity_t_exp}), we find that 
$t_0\lesssim t_{exp}$.
Since physical causality requires $t_{exp}<t_0$, we draw the qualitative
conclusion that $t_0\simeq t_{exp}$, which is consistent to within the uncertainties. Again, we caution that this derivation of $t_\mathrm{exp}$ relies on the assumption $v \propto t^{-0.22}$, which may not be valid for all SNe Ia.

@@

To examine the possibility of a SN-companion signature in the early
light curve of \abc, we employ the \citet{2010ApJ...708.1025K}
model and assume canonical values for the ejecta mass, $1.4\,\sm$, expansion velocity,
$10^{4}\,\textrm{km}\,\textrm{s}^{-1}$, and a constant opacity, 
$0.2\,\textrm{cm}^2\,\textrm{g}^{-1}$. We calculate the expected 
$g_\mathrm{PTF}$ brightness of an ejecta-companion collision at the 
distance of \abc\ behind a total redenning of $E(B-V) = 0.08 \, \mathrm{mag}$ using the parameterized equations in 
\citet{2012ApJ...749...18B}. If we assume the binary is aligned with the optimal orientation relative to the line of sight, a binary separation of $a \approx 2 \times 10^{13}\, \mathrm{cm}$ is needed to explain the initial detection of \abc, as shown in Figure~\ref{fig:SN-companion}. This model does not, however, match the $g_\mathrm{PTF}$ evolution for $t > t_0 + 1 \, \mathrm{d}$ (though it is possible that the SN photosphere dominates the companion-interaction signature at this phase). Furthermore, this model significantly overpredicts the observed flux in the \textit{Swift}/$UVW1$ filter $\sim$3.1~d after $t_0$. As such, we do not favor such a model as the explanation for the early emission from \abc.

@@

The Intermediate Palomar Transient Factory project is a scientific collaboration among the California Institute of Technology, Los Alamos National Laboratory, the University of Wisconsin, Milwaukee, the Oskar Klein Center, the Weizmann Institute of Science, the TANGO Program of the University System of Taiwan, and the Kavli Institute for the Physics and Mathematics of the Universe. This work was supported by the GROWTH project funded by the National Science Foundation under Grant No 1545949. Part of this research was carried out at the Jet Propulsion Laboratory, California Institute of Technology, under a contract with the NASA. This work makes use of observations from the LCO network. These results made use of the Discovery Channel Telescope at Lowell Observatory. Lowell is a private, non-profit institution dedicated to astrophysical research and public appreciation of astronomy and operates the DCT in partnership with Boston University, the University of Maryland, the University of Toledo, Northern Arizona University and Yale University.  The upgrade of the DeVeny optical spectrograph has been funded by a generous grant from John and Ginger Giovale.

@@

\software{\texttt{PTFIDE} \citep{Masci2017PASP}, \texttt{SALT2}, \texttt{SNID}, \texttt{lcogtsnpipe}, \texttt{UVOTSOURCE}}.
